---
title: "Reproducible Research Course Project 1"
author: "vlwhite10"
date: "September 3, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read In Libraries
```{r libraries, include=TRUE, warning=FALSE, results='hide'}
library(dplyr)
library(ggplot2)
```

# Read In Data
```{r data}
ActivityData<-read.csv("activity.csv")
```


# What is the mean number of steps taken per day?
```{r TotalSteps, results='hide'}
  #Sum the amount of steps each day
  ActivityData_by_Date<- ActivityData %>%
    group_by(factor(date)) %>%
    summarise(total.steps = sum(steps, na.rm=T))

  #Create Histogram of the Total Steps taken each Day
  ggplot(ActivityData_by_Date, aes(total.steps) ) + 
    geom_histogram(bins = 20) +
    labs(x = "Total Steps Each Day", y = "Count of Days", title="Histogram of Total Steps Each Day")
```

```{r MeanAndMedianSteps}
  meanSteps<-round(mean(ActivityData_by_Date$total.steps), digits = 0)
  medianSteps<-round(median(ActivityData_by_Date$total.steps), digits=0)
```
  
The mean amount of steps taken per day is `r meanSteps`. The median amount of steps taken per day is `r medianSteps`.

#What is the average daily activity pattern?
```{r AverageStepsPerInterval, results='hide'}
  #Average amount of steps in each interval
  ActivityData_by_Interval<- ActivityData %>%
    group_by(interval) %>%
    summarise(average.steps = mean(steps, na.rm=T))

  #Create Line graph for the average amount of steps taken in each 5-min interval
  ggplot(ActivityData_by_Interval, aes(y=average.steps, x=interval) ) + 
    geom_line() +
    labs(x = "5-min Interval", y = "Average Steps Taken", title="Average Steps Taken for each Interval")
  
  #Interval with maximum number of steps
  max_Steps<-ActivityData_by_Interval[which.max(ActivityData_by_Interval$average.steps),]

```
  
  The `r max_Steps$interval` interval contains the maximum amount of steps on average.
  
#Imputing Missing Values
```{R Count_missing_Values}
  #Count of missing values in data set
  Count_NA <- sum(is.na(ActivityData$steps))
```

There are `r Count_NA` intervals with missing step counts. 

```{R Replace_missing_values}
  #Identify Rows missing step values
  ActivityData_OnlyNA <- ActivityData[is.na(ActivityData$steps),]

  #Merge Intervals with missing step values to the interval average steps
  ActivityData_ReplacedNA <- merge(ActivityData_OnlyNA, ActivityData_by_Interval, by=c("interval"))
  ActivityData_ReplacedNA <- ActivityData_ReplacedNA[c("average.steps", "date", "interval")]
  names(ActivityData_ReplacedNA) <- c("steps", "date", "interval")
  
  #Data set without missing step values
  ActivityData_NoNA <- ActivityData[is.na(ActivityData$steps)==FALSE,]
  
  #Bind data with replaced NAs and without NAs
  ActivityData_NoNA <- rbind(ActivityData_NoNA, ActivityData_ReplacedNA)
  
  
  ActivityData_NoNA_by_Date<- ActivityData_NoNA %>%
  group_by(factor(date)) %>%
  summarise(total.steps = sum(steps, na.rm=T))
```

```{R Histogram with NA replaced with interval mean}

  #Create Histogram of the Total Steps taken each Day
  ggplot(ActivityData_NoNA_by_Date, aes(total.steps) ) + 
    geom_histogram(bins = 20) +
    labs(x = "Total Steps Each Day", y = "Count of Days", title="Histogram of Total Steps Each Day (NA values replaced with interval mean)")
```

```{R Mean and Medium of Dataset with NA replaced by interal mean}
  NoNAMean <- round(mean(ActivityData_NoNA_by_Date$total.steps), digit=0)
  NoNAMedian <- round(median(ActivityData_NoNA_by_Date$total.steps), digit = 0)
  
  MeanDiff <- NoNAMean - meanSteps
  MedianDiff <- NoNAMedian - medianSteps
```

The mean and median is shifted to the right when imputing the missing values with the mean of the interval. The mean shifts from `r meanSteps` to `r NoNAMean` a difference of `r MeanDiff`. The median shifts from `r medianSteps` to `r NoNAMedian` a difference of `r MedianDiff`. 

#Average Steps by Weekday versus Weekend

```{R NoNA data set with weekday and weekend identifier}
  ActivityData_NoNA$AsDate <- as.Date(ActivityData_NoNA$date, format = "%Y-%m-%d")
  
  ActivityData_NoNA$Weekday <- ifelse(weekdays(ActivityData_NoNA$AsDate) == 'Saturday' | weekdays(ActivityData_NoNA$AsDate) == 'Sunday', 'weekend', 'weekday')
  
  ActivityData_NoNA_by_Weekday<- ActivityData_NoNA %>%
    group_by(interval, Weekday) %>%
    summarise(average.steps = mean(steps, na.rm=T))
  
  ggplot(ActivityData_NoNA_by_Weekday, aes(interval, average.steps)) +
    geom_line()+ggtitle("Average Steps per Interval for Weekdays versus Weekends")+
    xlab("Interval") + ylab("Average Amount of Steps") +facet_grid(Weekday~.)
  
```

